Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-18 dev/notebooks/experiments/mitigation_strategies.qmd
cant get tables to work, charts will have to do

@@

@@

>>>>

### Bootstrap

```{julia}
n_bootstrap = 1000
using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
using DataFrames
df = DataFrame()
for (key, val) in results
    n_folds = length(val.experiment.recourse_systems)
    for fold in 1:n_folds
        for i in length(val.experiment.system_identifiers)
            rec_sys = val.experiment.recourse_systems[fold][i]
            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
            df_.model .= model_name
            df_.generator .= gen_name
            df_.fold .= fold
            df = vcat(df, df_)
        end
    end
end
df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
using RCall
save_path = joinpath(output_path, "bootstrap_synthetic.csv")
using CSV
CSV.write(save_path)
```

@@

### Bootstrap

```{julia}
n_bootstrap = 1000
using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
using DataFrames
df = DataFrame()
for (key, val) in results
    n_folds = length(val.experiment.recourse_systems)
    for fold in 1:n_folds
        for i in length(val.experiment.system_identifiers)
            rec_sys = val.experiment.recourse_systems[fold][i]
            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
            df_.model .= model_name
            df_.generator .= gen_name
            df_.fold .= fold
            df = vcat(df, df_)
        end
    end
end
df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
using RCall
save_path = joinpath(output_path, "bootstrap_latent.csv")
using CSV
CSV.write(save_path)
```

@@

### Bootstrap

```{julia}
n_bootstrap = 1000
using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
using DataFrames
df = DataFrame()
for (key, val) in results
    n_folds = length(val.experiment.recourse_systems)
    for fold in 1:n_folds
        for i in length(val.experiment.system_identifiers)
            rec_sys = val.experiment.recourse_systems[fold][i]
            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
            df_.model .= model_name
            df_.generator .= gen_name
            df_.fold .= fold
            df = vcat(df, df_)
        end
    end
end
df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
using RCall
save_path = joinpath(output_path, "bootstrap_real_world.csv")
using CSV
CSV.write(save_path)
```

### 2022-08-18 dev/notebooks/experiments/real_world.qmd
cant get tables to work, charts will have to do

>>>>

### Bootstrap

```{julia}
n_bootstrap = 1000
using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
using DataFrames
df = DataFrame()
for (key, val) in results
    n_folds = length(val.experiment.recourse_systems)
    for fold in 1:n_folds
        for i in length(val.experiment.system_identifiers)
            rec_sys = val.experiment.recourse_systems[fold][i]
            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
            df_.model .= model_name
            df_.generator .= gen_name
            df_.fold .= fold
            df = vcat(df, df_)
        end
    end
end
df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
using RCall
save_path = joinpath(output_path, "bootstrap.csv")
using CSV
CSV.write(save_path)
```

### 2022-08-18 dev/notebooks/experiments/synthetic.qmd
cant get tables to work, charts will have to do

### Tables

@tbl-results shows a summary of all results.

```{julia}
#| label: tbl-results
#| tbl-cap: "Summary of all results"
#| eval: true
#| results: asis
using AlgorithmicRecourseDynamics: kable
kable(results, [50]; format="html")
```

>>>>

### Bootstrap

```{julia}
n_bootstrap = 1000
using AlgorithmicRecourseDynamics.Evaluation: evaluate_system
using DataFrames
df = DataFrame()
for (key, val) in results
    n_folds = length(val.experiment.recourse_systems)
    for fold in 1:n_folds
        for i in length(val.experiment.system_identifiers)
            rec_sys = val.experiment.recourse_systems[fold][i]
            model_name, gen_name = collect(val.experiment.system_identifiers)[i]
            df_ = evaluate_system(rec_sys, val.experiment; n=n_bootstrap)
            df_.model .= model_name
            df_.generator .= gen_name
            df_.fold .= fold
            df = vcat(df, df_)
        end
    end
end
df = mapcols(x -> typeof(x) == Vector{Symbol} ? string.(x) : x, df)
using RCall
save_path = joinpath(output_path, "bootstrap.csv")
using CSV
CSV.write(save_path)
```

### 2022-08-18 dev/notebooks/generators/clap_roar_generator.qmd
sorted out deps and set up online appendix
uh
cant get tables to work, charts will have to do

---
title: ClapROAR Generator
jupyter: julia-1.7
---

@@

# `ClapROARGenerator`

@@

```{julia}
plts = []
for (name,ce) ∈ counterfactuals
    plt = plot(ce, plot_up_to=minimum([T,T_]), title=name)
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(800,300), layout=(1,2))
```

```{julia}

```

>>>>

## ClapROAR Generator

@@

@@

```{julia}
plts = []
for (name,ce) ∈ counterfactuals
    plt = plot(ce, plot_up_to=minimum([T,T_]), title=name)
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(800,300), layout=(1,2))
display(plt)
```

### 2022-08-18 paper/sections/empirical.rmd
bleh
cant get tables to work, charts will have to do

Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary. Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019towards, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concered with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

@@

Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 2,500 individuals each. We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.

>>>>

Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary. Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019towards, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been omitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

@@

Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 2,500 individuals each. We have also standardized all explanatory features since our chosen classifiers are sensitive to scale.

### 2022-08-18 paper/sections/empirical_2.rmd
cant get tables to work, charts will have to do

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.]. Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn). The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

>>>>

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.]. Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn). The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

### 2022-08-18 paper/sections/introduction.rmd
cant get tables to work, charts will have to do

::: {.example #student name="Student Admission"}
Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a university in their student admission process. Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more liekly to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
:::

>>>>

::: {.example #student name="Student Admission"}
Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a university in their student admission process. Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
:::

### 2022-08-18 paper/sections/mitigation.rmd
cant get tables to work, charts will have to do

Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrent our attention, in this section we introduce and test a number of potential mitigation strategies. All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process. This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section. But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and - perhaps more importantly - their performance hinges on the performance of said generative models. Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.

@@

Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Graviational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].

>>>>

Having established in the previous section that endogenous macrodynamics in AR are substantial enough to warrant our attention, in this section we introduce and test a number of potential mitigation strategies. All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process. This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section. But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and - perhaps more importantly - their performance hinges on the performance of said generative models. Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.

@@

Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].

### 2022-08-18 paper/sections/related.rmd
cant get tables to work, charts will have to do

Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017. An October 2020 survey by Karimi et al. layed out 60 algorithms that have been proposed since 2014 [@karimi2020survey]. Another survey published around the same time by Verma et al. described 29 algorithms [@verma2020counterfactual]. Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met. Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs. To name but a few, we are interested in generating counterfactuals that close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and if possible causally founded [@karimi2021algorithmic].

@@

Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia. Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits. In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: [`AlgorithmicRecourseDynamics.jl`]((https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md))^[The code is available from the following anonymized Github repository: [https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).]. The core concepts implemented in that package reflect what is presented in Section \@ref(method-2) of this paper.

>>>>

Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017. An October 2020 survey by Karimi et al. laid out 60 algorithms that have been proposed since 2014 [@karimi2020survey]. Another survey published around the same time by Verma et al. described 29 algorithms [@verma2020counterfactual]. Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met. Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs. To name but a few, we are interested in generating counterfactuals that close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and if possible causally founded [@karimi2021algorithmic].

@@

Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic approaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, `CounterfactualExplanations.jl` [@altmeyer2022CounterfactualExplanations] provides an extensible, fast and language-agnostic implementation in Julia. Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits. In particular, we have built a framework on top of `CounterfactualExplanations.jl` that extends the functionality from static benchmarks to simulation experiments: [`AlgorithmicRecourseDynamics.jl`]((https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md))^[The code is available from the following anonymized Github repository: [https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md).]. The core concepts implemented in that package reflect what is presented in Section \@ref(method-2) of this paper.
